---
title: "Project Proposal"
author: "Nathan Bush, Lilah DuBoff, Syed Huma Shah, and Atreya Tadepalli"
subtitle: "Due November 17 at 11:59pm"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

#### Load Packages

```{r load-packages, message = FALSE, warning = FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com"))
library(tidyverse)
library(dplyr)
library(ggplot2)
install.packages("corrplot")
library(corrplot)


rm(list = ls())
```

# Dataset 1

**Data source:** Mendeley
Data, [https://data.mendeley.com/datasets/wj9rwkp9c2/1](https://www.google.com/url?q=https%3A%2F%2Fdata.mendeley.com%2Fdatasets%2Fwj9rwkp9c2%2F1)

**Brief description:** The data were collected from an Iraqi laboratory
within the Medical City Hospital, the Specialized Center for
Endocrinology, and the Diabetes-Al-Kindy Teaching Hospital. The dataset
is composed of patient files, inlcuding medical information, laboratory
analysis, and diagnoses.

**Long description:** This dataset contains information about
individuals who are classified as not having diabetes, are prediabetic,
or have been diagnosed with diabetes. Variables include:

1.  Patient ID
2.  Blood Sugar Level
3.  Age
4.  Gender
5.  Creatinine ratio(Cr)
6.  Body Mass Index (BMI)
7.  Urea
8.  Cholesterol (Chol)
9.  Fasting lipid profile including total, LDL, VLDL, Triglycerides(TG)
    and HDL Cholesterol
10. HBA1C
11. Class (the patient's diabetes disease class may be Diabetic,
    Non-Diabetic, or Pre-Diabetic).

The dataset allows for an investigation into the factors that contribute
to the risk of developing diabetes, and provides an opportunity to
uncover patterns and relationships that could inform both clinical
practice and public health interventions. The variables can be
classified into three types: biochemical markers (blood sugar, urea,
creatinine), lipid profiles, and HbA1c levels. Together, these variables
are essential for understanding the metabolic profile of individuals
across different diabetes classifications. The age and gender categories
can help to identify the most-at risk groups.

\newpage

**Research question 1:** To what extent do demographic factors, such as
age and gender, influence a patient’s likelihood of being classified as
diabetic, pre-diabetic, or non-diabetic? Additionally, how do these
demographic characteristics interact with other clinical variables, such
as BMI, blood sugar levels, or lipid profiles, in determining diabetes
risk and progression?

**Outcome Variable:** CLASS (Diabetes status of patient)

**Type:** Ordinal Categorical Variable

**Predictor Variables:** Age, Gender, potentially other interaction
variables

**Inference Goal:** This question seeks to determine if either age or
gender are statistically significant variables when predicting the
diabetes class of a patient, and if age interacting with other clinical
variables has an impact on classification.

**Research question 2:** Do health information statistics like blood
sugar level, cholesterol, or lipids profiles have an impact on whether a
patient is classified as diabetic, pre-diabetic, or non-diabetic?

**Outcome Variable Name:** *CLASS* (Diabetes status of patient)

**Type:** Ordinal Categorical Variable

**Predictor Variables:**

*Urea* - Urea amount (Continuous, mmol/L)

*Chol* - Cholesterol measurement (Continuous, mg/dL)

*BMI* - Body Mass Index (Continuous)

*HBA1C* - Hemoglobin A1C; long-term blood sugar levels (Continuous, %)

*Cr* - Creatinine ratio (Continuous, mg/g)

*TG* - Triglycerides (Continuous, mg/dL)

*HDL* - High-density Lipoprotein Cholesterol (Continuous, mg/dL)

*VLDL* - Very Low-density Lipoprotein Cholesterol (Continuous, mg/dL)

*LDL* - Low-density Lipoprotein Cholesterol (Continuous, mg/dL)

**Description:** This question seeks to determine if the diabetes class
of a patient can be accurately predicted using the health indicators
available\

\newpage

***Research Question 1 EDA and Plots***

```{r}
diabetes_df <- read.csv("https://raw.githubusercontent.com/lilah-duboff/Stats_Final_Project/refs/heads/main/Dataset%20of%20Diabetes%20.csv")

glimpse(diabetes_df)
```

```{r}
diabetes_df |> count(CLASS)
diabetes_df |> count(Gender)
```

As we can see from these counts tables, there are a few issues with how
the data was input into the dataset. Notably, there are duplicate labels
for how the diabetes classification was input - this is likely due to
white-space, so in the next chunk, we've removed it, and checked the
table counts again. In the gender table, there is one entry where a
female patient was denoted with a lowercase f, instead of a capital F.
We have changed this as well.

```{r}
# Remove extra spaces and standardize to uppercase
diabetes_df$CLASS <- toupper(trimws(diabetes_df$CLASS))
diabetes_df$Gender[diabetes_df$Gender == "f"] <- "F"

diabetes_df |> count(CLASS)
diabetes_df |> count(Gender)
```

We checked the counts again, and all seems to be consistent!

\newpage

***Demographics Plots***

```{r}
ggplot(diabetes_df, aes(x = CLASS)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(
    title = "Counts of Diabetes Diagnoses",
    x = "Category",
    y = "Count") 
```

This bar graph shows the counts of patients who were identified as
either non-diabetic (N), Pre-Diabetic (P), or Diabetic (Y). We can see
that the vast majority of patients in the dataset were classified as
diabetic, which could potentially influence analysis.

\newpage

```{r}
ggplot(data = diabetes_df, aes(x = CLASS, y = AGE, fill = CLASS)) + 
  geom_boxplot() +
  ggtitle("Age and Diabetes Class") +
  theme_minimal() + 
  scale_fill_manual(values = c("Y" = "red", "P" = "yellow", "N" = "green"))


ggplot(data = diabetes_df, aes(x = CLASS, y = BMI, fill = CLASS)) + 
  geom_boxplot() +
  ggtitle("BMI and Diabetes Class") +
  theme_minimal() + 
  scale_fill_manual(values = c("Y" = "red", "P" = "yellow", "N" = "green"))
```

The first box plot illustrates the relationship between age and diabetes
class (Y = Yes, P = Pre-diabetes, N = No). We can see that the median
age increases progressively from the "No" (N) category to the
"Pre-diabetes" (P) and then to the "Diabetes" (Y) category. This trend
suggests that individuals with a diabetes diagnosis tend to be older on
average compared to those without diabetes or in the pre-diabetes
category. The distribution in the "Y" - or "Diabetes" - class shows a
wider range of ages, indicating more variability in the ages of people
diagnosed with diabetes.

The second box plot shows the distribution of Body Mass Index (BMI)
across the same diabetes categories. Here, we see a similar trend where
the median BMI increases from the "No" (N) to "Pre-diabetes" (P) and
then to the "Diabetes" (Y) category. Individuals with diabetes (Y) have
a notably higher median BMI compared to those in the pre-diabetes or
non-diabetes groups. The wider spread in the "Diabetes" category
suggests that BMI varies more among individuals with diabetes,
highlighting the potential link between obesity and the progression to
diabetes.

\newpage

***Clinical Measures Plots***

```{r}
ggplot(data = diabetes_df, aes(x = CLASS, y = HbA1c, fill = CLASS)) +
  geom_boxplot() +
  ggtitle("Blood Sugar Level and Diabetes Class") +
  theme_minimal() + 
  scale_fill_manual(values = c("Y" = "red", "P" = "yellow", "N" = "green"))


ggplot(data = diabetes_df, aes(x = CLASS, y = Chol, fill = CLASS)) +
  geom_boxplot() +
  ggtitle("Cholesterol and Diabetes Class") +
  theme_minimal() + 
  scale_fill_manual(values = c("Y" = "red", "P" = "yellow", "N" = "green"))

```

The first box plot shows the relationship between blood sugar levels
(HbA1c) and diabetes class. It is evident that there is a dramatic
difference in median HbA1c levels from the increases from the
"Non-Diabetic" (N) category to the "Diabetes" (Y) category. This is
expected, as medical knowledge states that individuals with higher blood
sugar levels are more likely to be at risk for pre-diabetes or diabetes.
The data indicates that blood sugar levels are higher and more variable
among those diagnosed with diabetes.

In contrast, the second box plot shows the distribution of cholesterol
levels across the same diabetes categories. Here, we do not observe a
clear trend in cholesterol levels between the "No" (N), "Pre-diabetes"
(P), and "Diabetes" (Y) groups. The median cholesterol levels remain
relatively stable across all categories, with no significant increase as
the diabetes class changes. This suggests that cholesterol may not be as
closely related to the classification of diabetes as blood sugar levels
(HbA1c), or that its role in diabetes risk may be more complex or
influenced by other factors.

\newpage

EDA Inquiry for Research Question 2: How correlated are the predictor
variables (ex. blood sugar level, cholesterol) that may be utilized to
predict the diabetes class of a patient?

-   This may serve to better understand the multicollinearity included
    within the model, and how to remove variables to strengthen the
    statistical interpretation of the remaining predictor variables
    included within the model.

```{r}
df_corr=cor(diabetes_df[sapply(diabetes_df,is.numeric)])
corrplot(df_corr)
```

From this correlation matrix of the numerical predictor variables, it
appears that there will not be major issues of multicollinearity when
preparing the model. The strongest correlation appears to be between age
and BMI, or Urea and Creatinine level. One of either of these variables
may be excluded from the regression, depending on testing.

\newpage

# Dataset 2

**Data source:**

https://catalog.data.gov/dataset/crime-data-from-2020-to-present

**Brief description:**

This data set reflects incidents of crime in the City of Los Angeles
dating from 2020 to 2023. This data is transcribed from original crime
reports that are typed on paper and therefore there may be some
inaccuracies within the data.

This code book describes the data in more depth:
https://data.lacity.org/Public-Safety/Crime-Data-from-2020-to-Present/2nrs-mtv8/about_data

**Long description**

The City of Los Angeles provides a crime dataset from 2020 to the 2023,
covering incidents reported to the Los Angeles Police Department (LAPD).
Here’s a breakdown of key columns in the dataset:

1\. DR_NO: Unique identifier assigned to each crime report, used to
track individual cases.

2\. Date Rptd: Date the incident was officially reported to the police.

3\. DATE OCC: Date when the crime actually occurred.

4\. TIME OCC: Time the incident occurred, which allows for time-of-day
analysis.

5\. AREA: Code representing the geographical area of Los Angeles where
the incident took place.

6\. AREA NAME: Name of the area corresponding to the AREA code,
providing a more human-readable location.

7\. Rpt Dist No: Reporting district number within the LAPD, which is a
more specific geographical indicator within the area.

8\. Part 1-2: Crime classification indicator distinguishing between Part
1 and Part 2 crimes, which helps in crime severity categorization.

9\. Crm Cd: Crime code representing a specific type of crime.

10\. Crm Cd Desc: Description of the crime type (e.g., robbery,
assault), giving context to the "Crm Cd" column.

11\. Mocodes: Modus operandi codes that describe the method or behavior
pattern of the suspect.

12\. Vict Age: Age of the victim, which can be used for demographic
analysis.

13\. Vict Sex: Gender of the victim, another demographic detail.

14\. Vict Descent: Ethnic background or descent of the victim.

15\. Premis Cd: Premise code indicating the type of location where the
crime occurred.

16\. Premis Desc: Description of the premise (e.g., street, residence),
helping to understand crime locations.

17\. Weapon Used Cd: Code indicating if a weapon was used in the crime,
which can be used to assess weapon involvement trends.

18\. Weapon Desc: Description of the weapon, if applicable, providing
details on the weapon type.

19\. Status: Code indicating the current status of the case (e.g., open,
closed).

20\. Status Desc: Description of the case status, complementing the
"Status" code with a text explanation.

21\. Crm Cd 1-4: Additional crime codes, capturing cases where multiple
types of crimes were involved in a single incident. 22. LOCATION:
General location description of the crime.

23\. Cross Street: Cross street information for more precise location
data. 24. LAT: Latitude coordinate of the crime location, useful for
mapping. 25. LON: Longitude coordinate of the crime location, also
useful for mapping.

This dataset allows for comprehensive analysis of crime trends in Los
Angeles, with potential insights into crime types, locations, times,
demographics of victims, weapon involvement, and case status. The
dataset is valuable for identifying patterns, conducting demographic
analysis, and mapping geographical hotspots of crime.

**Research question 1:**

**Research Question:** What is the relationship between the severity of
reported crimes and their spatiotemporal distribution in L.A? More
specifically how do the frequencies of Part 1 and Part 2 violent crimes
vary across different geographical areas of Los Angeles and the time of
the day between 2020 and 2023?

**Outcome Variable:** *Part 1-2* (binary variable indicating crime
seriousness, with Part 1 crimes generally more serious than Part 2).

**Predictor Variables:** *LAT* and *LON* (location coordinates), *AREA*
(area code), and *TIME OCC* (time of day). Maybe others too.

**Inference Goal:** This question seeks to determine if more serious
crimes are more prevalent in certain areas and at specific times.

**Research question 2:**

**Research Question:** How do victim demographics (age, sex, descent)
influence the likelihood of being involved in the most common
crimes—vehicle theft, simple assault (battery), burglary from vehicle,
theft of identity, and felony vandalism—in Los Angeles between 2020 and
2023, and how do these patterns vary across different geographical
areas?

**Outcome Variable Name:** `Crm.Cd.Desc` (Crime Code Description)\
**Type:** Categorical Variable (Nominal)\
**Description:**\
The outcome variable for this research question is the **type of crime
committed**, specifically focusing on the top five most common crimes in
Los Angeles between 2020 and 2023. These crimes are:

1.  **Vehicle - Stolen**

2.  **Battery - Simple Assault**

3.  **Burglary From Vehicle**

4.  **Theft of Identity**

5.  **Vandalism - Felony (\$400 & Over, All Church Vandalisms)**

This variable represents the specific crime associated with each
incident report in the dataset. It is a nominal categorical variable
because the categories (crime types) are names without an inherent order
or ranking.

**Predictor Variables:** - Vict.Age - Vict.Sex - Vict.Descent -
AREA.NAME

**Inference Goal:** This question aims to analyze the relationship
between victim demographics and the likelihood of being involved in the
five most common crimes in Los Angeles. It also seeks to investigate how
these patterns differ across various geographical areas within the city.
By focusing on these specific crimes and demographic factors, we can
identify potential vulnerabilities and trends in victimization across
different population groups and locations.

\newpage

**Load the data and provide a `glimpse()`**:

```{r data-2, warning=FALSE}

data <- read.csv("https://github.com/lilah-duboff/Stats_Final_Project/raw/refs/heads/main/Crime_Data_from_2020_to_Present.csv?download=")

glimpse(data)
```

**Exploratory Plots:** ***Research question 1 plots:***

```{r plot-2_1, echo=FALSE, size= 10 }
library(ggplot2)

# Assuming 'data' is your dataframe and '' is your binary variable
ggplot(data, aes(x = factor(Part.1.2))) +
  geom_bar(fill = "steelblue") +
  labs(x = "Crime type", y = "Count",title = "Dist of Part 1(serious crimes) & Part 2(less serious crimes)") +
  theme_minimal() +
  scale_x_discrete(labels = c("1" = "serious crimes", "2" = "less serious crimes"))
```

Distribution of Part 1 (Serious Crimes) & Part 2 (Less Serious Crimes)
This bar chart categorizes crimes into Part 1 (serious crimes) and Part
2 (less serious crimes) and shows the frequency of each category. The
two bars illustrate the relative proportions of serious versus less
serious crimes, providing insight into the overall severity distribution
within the dataset. EDA Insights: Part 1 crimes have a higher frequency
than Part 2 crimes, indicating that serious crimes make up a larger
portion of the reported incidents. The more severe crimes have about
591,254 observations and less severe crimes are 399,039. This
distribution helps understand the nature of crime severity in Los
Angeles, which may influence policing or resource allocation.

```{r plot-2_2, echo=FALSE }


# Create the crime area summary and recode the Part.1.2 values
crime_area_summary <- data %>%
  mutate(Part.1.2 = as.character(Part.1.2)) %>%  # Convert to character first
  mutate(Part.1.2 = fct_recode(Part.1.2,
    "Part 1 (serious crimes)" = "1",
    "Part 2 (less serious crimes)" = "2"
  )) %>%
  group_by(AREA.NAME, Part.1.2) %>%
  summarise(Frequency = n(), .groups = "drop")

# Create the heatmap with updated labels
ggplot(crime_area_summary, aes(x = Part.1.2, y = AREA.NAME, fill = Frequency)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red") +
  labs(title = "Frequency of Crimes by Area and Severity",
       x = "Crime Severity",
       y = "Area Name",
       fill = "Frequency") +
  theme_minimal() 

```

Frequency of Crime Severity( more severe and less severe) Description:
This heatmap shows the frequency of the top five crimes across various
geographical areas. Each cell's color intensity represents the count of
a particular crime type within an area, with darker shades indicating
higher frequencies. The y-axis lists the areas, while the x-axis lists
the severity of crime( binary: less severe and more severe), allowing
for a spatial view of crime distribution. EDA Insights: Areas with
darker cells have higher crime counts, highlighting regions with
potentially higher crime rates for each type. Certain areas like
"Pacific", "Central","77 Street", etc. have a consistently high
frequency for more severe crimes, indicating general high-crime zones.
Areas with high frequencies for severity of the crime could benefit from
targeted crime prevention programs or policing strategies. \newpage
***Research question 2 plots:***

```{r plot-2_3, echo=FALSE }
# Filter for the top five crimes
top_crimes <- c("VEHICLE - STOLEN", "BATTERY - SIMPLE ASSAULT",
                "BURGLARY FROM VEHICLE", "THEFT OF IDENTITY",
                "VANDALISM - FELONY ($400 & OVER, ALL CHURCH VANDALISMS)")
filtered_data <- data %>%
  filter(Crm.Cd.Desc %in% top_crimes)
# 1. Summary of the outcome variable
part_1_2_summary <- table(data$Part.1.2)

# Assuming 'data' is your dataframe and '' is your binary variable
ggplot(filtered_data, aes(x = factor(Crm.Cd.Desc))) +
  geom_bar(fill = "steelblue") +
  labs(x = "Crime Name", y = "Count",title = "Distribution of Top Five Crimes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 15))
```

Distribution of Top Five Crimes Description: This bar chart shows the
frequency of the top five most common crimes in Los Angeles. Each bar
represents a specific crime type, with the height indicating the total
count of occurrences. This visualization provides an overview of the
relative prevalence of each crime, helping to identify which types are
most frequent in the dataset. EDA Insights: The chart clearly shows that
"Vehicle - Stolen" has the highest frequency among the top five crimes,
followed closely by "Battery - Simple Assault." This distribution allows
for quick comparison across crime types, highlighting which crimes are
more common. Observing high counts for specific crimes could indicate
priority areas for law enforcement or community awareness programs.

```{r plot-2_4,fig.width=8, fig.height=7, echo=FALSE}
# Aggregate the data
crime_area_summary <- filtered_data %>%
  group_by(AREA.NAME, Crm.Cd.Desc) %>%
  summarise(Frequency = n())

# Create the heatmap with improved x-axis labels
ggplot(crime_area_summary, aes(x = Crm.Cd.Desc, y = AREA.NAME, fill = Frequency)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red") +
  labs(title = "Frequency of Top 5 Crimes by Area",
       x = "Crime Type",
       y = "Area Name",
       fill = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 15))


```

Frequency of Top 5 Crimes by Area Description: This heatmap shows the
frequency of the top five crimes across various geographical areas. Each
cell's color intensity represents the count of a particular crime type
within an area, with darker shades indicating higher frequencies. The
y-axis lists the areas, while the x-axis lists the crime types, allowing
for a spatial view of crime distribution. EDA Insights: Areas with
darker cells have higher crime counts, highlighting regions with
potentially higher crime rates for each type. Certain areas like
"Pacific", "Central","77 Street", etc. have a consistently high
frequency for crime types like "Vandalism" and "Central" area has
highest " Burglary from Vehicle" crime type cases. Areas with high
frequencies for specific crime types could benefit from targeted crime
prevention programs or policing strategies.

```{r plot-2_5 ,fig.width=11,fig.height=5 , echo=FALSE }
# Filter for valid Vict Age and the top five crimes
filtered_data <- filtered_data %>%
  mutate(Vict.Age = as.numeric(Vict.Age)) %>%
  filter(!is.na(Vict.Age) & Vict.Age > 0 & Vict.Age < 120, Vict.Sex %in% c('M', 'F'))

# Create the plot
ggplot(filtered_data, aes(x = Vict.Age, fill = Vict.Sex)) +
  geom_histogram(binwidth = 5, alpha = 0.7, position = "identity") +
  facet_grid(Vict.Sex ~ Crm.Cd.Desc) +
  labs(title = "Age Distribution of Victims by Crime Type and Sex",
       x = "Victim Age",
       y = "Number of Victims",
       fill = "Victim Sex") +
  theme_minimal()

```

Age Distribution of Victims by Crime Type and Sex Description: This
faceted plot shows the age distribution of victims across different
crime types, separated by victim sex (female and male). Each facet
represents a unique combination of crime type and sex, with the x-axis
showing victim age and the y-axis showing the count. This plot reveals
age-based patterns in victimization for each crime type, broken down by
gender. EDA Insights: Certain age groups may have higher victimization
rates for specific crimes, suggesting patterns of vulnerability related
to age. Comparing male and female distributions within each crime type
can reveal gender-based differences in victimization, potentially
indicating targeted or biased victimization. There is a noticeably
higher count of female victims across multiple crime types, particularly
in simple assault and identity theft. Male victim counts are generally
lower than females for most crimes; however, certain crimes, such as
vehicle theft and burglary from vehicles, show a more balanced
distribution between genders. This could suggest that these crime types
are less influenced by the victim's gender. Across most crimes, young
adults appear to have the highest victimization rates, particularly
visible in crimes like simple assault and identity theft. This age group
might be more exposed to environments or activities associated with
these crimes. Different crimes show unique distributions by age and
gender, which can guide more focused safety or awareness campaigns for
specific demographic groups.
